---
title: "p8105_hw2_os2424"
author: "Ou Sha"
output: github_document
date: "2023-10-04"
---

```{r setup, include=FALSE}
library(tidyverse)
library(lubridate)
library(dplyr)
library(readxl)
```

# Problem 1
```{r}
# import data set
pols <- read.csv("./fivethirtyeight_datasets/pols-month.csv")
# clean pols-month data
pols <- janitor::clean_names(pols)
pols <- pols |>
  # break up variable mon into integer variables
  separate(mon, c("year", "month", "day"), "-")|>
  #create president variable
  mutate(president = if_else(prez_gop== 0,'dem','gop'),
         # replace month number with month name
         month = month.name[as.numeric(month)])|>
  # remove prez_dem, prez_gop, day variables
  dplyr::select(-prez_gop, -prez_dem, -day)
```

```{r}
# import data set
snp <- read.csv("./fivethirtyeight_datasets/snp.csv")
# clean snp data
snp <- janitor::clean_names(snp)
snp <- snp |>
  # break up variable date into integer variables
  separate(date, c("month", "day", "year"), "/")|>
  # organize year
  mutate(year = if_else(year<= 15, paste("20", year, sep=""), paste("19", year, sep="")),
         # replace month number with month name
         month = month.name[as.numeric(month)])|>
  # arrange by year and month
  arrange(year, month)|>
  # make year and month leading columns
  dplyr::select(year, month, everything(), -day)
```

```{r}
# import data set
unemployment <- read.csv("./fivethirtyeight_datasets/unemployment.csv")
# switch from “wide” to “long” format
unemployment <- unemployment|>
  pivot_longer(
    Jan:Dec,
    names_to = "month", 
    values_to = "percent")|>
  # make variable type the same to merge
  mutate(year = as.character(Year))|>
  dplyr::select(year, month, percent)
# clean data 
unemployment <- janitor::clean_names(unemployment)
# merge snp into pols, and merge unemployment into the result
merged <- left_join(pols, snp, by = join_by(year, month))
merged <- left_join(merged, unemployment, by = join_by(year, month))
```
In pols data set, there are `r nrow(pols)` observations and `r ncol(pols)` variables. It includes the number of republican and democratic governors, senators, and representatives on the associated date. Whether the president was democratic or republican onthe associated date. \
In snp data set, there are `r nrow(snp)` observations and `r ncol(snp)` variables. It includes the closing values of the S&P stock index on the associated date ranging from `r range(snp$year)[1]` to `r range(snp$year)[2]`.\
In unemployment data set, there are `r nrow(unemployment)` observations and `r ncol(unemployment)` variables. It indicates the percentage of unemployment every month ranging from `r range(unemployment$year)[1]` to `r range(unemployment$year)[2]`.

# Problem 2
```{r}
# import Mr. Trash wheel data set (skip figure)
trash_wheel <- read_excel("./202309 Trash Wheel Collection Data.xlsx", "Mr. Trash Wheel", "A2:N586")
# use reasonable variables and clean data
trash_wheel <- janitor::clean_names(trash_wheel)
trash_wheel <- trash_wheel |>
  dplyr::select(-year, -month)|>
  # add new homes_powered variable
  mutate(homes_powered = weight_tons * 500/30)
```

```{r}
# import Professor Trash wheel data set (skip figure)
pf_trash <- read_excel("./202309 Trash Wheel Collection Data.xlsx", "Professor Trash Wheel", "A2:M108")
# use reasonable variables and clean data
pf_trash <- janitor::clean_names(pf_trash)
pf_trash <- pf_trash |>
  dplyr::select(-year, -month)|>
  # add = homes_powered variable
  mutate(homes_powered = weight_tons * 500/30)
```

```{r}
# import Cwynnda Trash wheel data set (skip figure)
G_trash <- read_excel("./202309 Trash Wheel Collection Data.xlsx", "Gwynnda Trash Wheel", "A2:L157")
# use reasonable variables and clean data
G_trash <- janitor::clean_names(G_trash)
G_trash <- G_trash |>
  dplyr::select(-year, -month)|>
  # add new homes_powered variable
  mutate(homes_powered = weight_tons * 500/30)
```

```{r}
# add new variable to track trash wheel type
G_trash <- G_trash |>
  mutate(type = "Gwynnda Trash Wheel")
pf_trash <- pf_trash |>
  mutate(type = "Professor Trash Wheel")
trash_wheel <- trash_wheel |>
  mutate(type = "Mr. Trash Wheel")
# combine data
combined_trash <- bind_rows(trash_wheel, pf_trash, G_trash)
combined_trash <- combined_trash |>
  #arrange by date
  arrange(date)|>
  # separate date variable
  separate(date, c("year", "month", "day" ), "-")|>
  # organize month variable
  mutate(month= month.name[as.numeric(month)])
```
In Mr. Trash Wheel data set, there are `r nrow(trash_wheel)` observations and `r ncol(trash_wheel)` variables. In Professor Trash Wheel data set, there are `r nrow(pf_trash)` observations and `r ncol(pf_trash)` variables. In Gwynnda Trash Wheel data set, there are `r nrow(G_trash)` observations and `r ncol(G_trash)` variables.\
In combined data, there are `r nrow(combined_trash)` observations and `r ncol(combined_trash)` variables. Key variable `homes_powered` is number of homes powered by the trash, `type` is the source of trash wheel.

```{r}
# total weight of trash by Professor Trash Wheel
weight_trash <- sum(pull(pf_trash, weight_tons))
# total number of cigarette butts by Gwynnda in July 2021
G_July <- combined_trash |>
  filter(type=="Gwynnda Trash Wheel"& month=="July" & year=="2021")
cig <- sum(pull(G_July, cigarette_butts))
```
The total weight of trash collected by Professor Trash Wheel is `r weight_trash`.\
The total number of cigarette butts collected by Gwynnda in July of 2021 is `r cig`.


# Problem 3
```{r}
# import MCI basline data
mci_baseline <- read.csv("./data_mci/MCI_baseline.csv", skip = 1)
# clean data
mci_baseline <- janitor::clean_names(mci_baseline)
mci_baseline <- mci_baseline |>
  # Ensure that sex and APOE4 carrier status are appropriate encoded
  mutate(sex = if_else(sex==0, "female", "male"),
         apoe4 = if_else(apoe4==0, "non-carrier", "carrier"))
mci_developed <- mci_baseline |>
  # remove any participants who do not meet the stated inclusion criteria
    mutate(age_at_onset=if_else(age_at_onset==".", NA, age_at_onset))|>
    filter(current_age<age_at_onset | is.na(age_at_onset))
```
The important steps in the import process are to skip the first row in the data set, and understand what each number entry for variable sex and apoex represents. The data set has `r nrow(mci_baseline)` observations and `r ncol(mci_baseline)` variables. `age_on_set` is the age at the onset of MCI. 

```{r}
# find how many participants were recruited
recruited <- nrow(mci_developed)
# find how many participants develop MCI
Participants_mci <- nrow(filter(mci_developed, !is.na(age_at_onset)))
# find average baseline age
average_age <- mean(pull(mci_developed, current_age))
# find proportion of women in the study are apoe4 carriers
proportion_w <- nrow(filter(mci_developed, sex=="female" & apoe4=="carrier"))/ 
  nrow(filter(mci_developed, sex=="female"))
```
`r recruited` participants were recruited.\
`r Participants_mci` of these how many develop MCI.\
The average baseline age is `r average_age`.\
Proportion of women in the study are APOE4 carriers is `r proportion_w`

```{r}
# import MCI amyloid data
mci_amyloid <- read.csv("./data_mci/mci_amyloid.csv", skip = 1)
# clean data
mci_amyloid <- janitor::clean_names(mci_amyloid)
mci_amyloid<- mci_amyloid |>
  # Ensure that sex and APOE4 carrier status are appropriate encoded
  rename(id=study_id, time_0=baseline)|>
  # make it to long format
  pivot_longer(
    time_0:time_8,
    names_to = "time", 
    values_to = "ratio")
```
The important steps in the import process are to skip the first row in the data set, and rename the study_id variable for further combination, and rename basline varible to time_0 to make the data set to long format. `baseline`(`time_0`) is the time elapsed since the study baseine to the visit where biomarker amyloid was measured. 

```{r}
#  Check whether some participants appear in only the amyloid datasets
amyloid_only <- anti_join(mci_amyloid, mci_developed, by="id")
#  Check whether some participants appear in only the baseline datasets
baseline_only <- anti_join(mci_developed, mci_amyloid,by="id")
```
`r nrow(amyloid_only)/5` participants appear in only the amyloid datasets, and their IDs: `r unique(pull(amyloid_only, "id"))`.\
`r nrow(baseline_only) ` participants appear in only the baseline datasets, and their IDs: `r pull(baseline_only, "id")`.

```{r}
# combine the demographic and biomarker datasets
combined_mci <- inner_join(mci_developed, mci_amyloid, by="id")
# export combined data
write.csv(combined_mci, file = "./data_mci/combined data.csv", row.names = TRUE)
```
The combined data set has `r nrow(combined_mci)` rows and `r ncol(combined_mci)` columns. Since each id has 5 rows of data, there are `r nrow(combined_mci)/5` observed individuals, and `r nrow(combined_mci)` observations in total. Age, year of education, apoe4 carrier or not, age at onset, ratio of years of measured amyloid ratio, and tears elapsed since the study baseline to the visit where amyloid was measured are included in the data set.