---
title: "p8105_hw1_os2424"
author: "Ou Sha"
output: github_document
date: "2023-10-04"
---

```{r setup, include=FALSE}
library(tidyverse)
library(lubridate)
library(dplyr)
library(readxl)
```

# Problem 1
```{r}
# import data set
pols <- read.csv("./fivethirtyeight_datasets/pols-month.csv")
# clean pols-month data
pols <- janitor::clean_names(pols)
pols <- pols|>
  # break up variable mon into integer variables
  separate(mon, c("year", "month", "day"), "-")|>
  #create president variable
  mutate(president = if_else(prez_gop == 0,'dem','gop'),
         # replace month number with month name
         month = month.name[as.numeric(month)]) |>
  # remove prez_dem, prez_gop, day variables
  dplyr::select(-prez_gop, -prez_dem, -day)
```

```{r}
# import data set
snp <- read.csv("./fivethirtyeight_datasets/snp.csv")
# clean snp data
snp <- janitor::clean_names(snp)
snp <- snp|>
  # break up variable date into integer variables
  separate(date, c("month", "day", "year"), "/")|>
  # organize year
  mutate(year = if_else(year <= 15, paste("20", year, sep=""), paste("19", year, sep="")),
         # replace month number with month name
         month = month.name[as.numeric(month)])|>
  # arrange by year and month
  arrange(year, month)|>
  # make year and month leading columns
  dplyr::select(year, month, everything(), -day)
```

```{r}
# import data set
unemployment <- read.csv("./fivethirtyeight_datasets/unemployment.csv")
# switch from “wide” to “long” format
unemployment <- unemployment|>
  pivot_longer(
    Jan:Dec,
    names_to = "month", 
    values_to = "percent")|>
  # make variable type the same to merge
  mutate(year = as.character(Year))|>
  dplyr::select(year, month, percent)
# clean data 
unemployment <- janitor::clean_names(unemployment)
# merge snp into pols, and merge unemployment into the result
merged <- left_join(pols, snp, by = join_by(year, month))
merged <- left_join(merged, unemployment, by = join_by(year, month))
```
In pols data set, there are `r nrow(pols)` observations and `r ncol(pols)` variables. It includes the number of republican and democratic governors, senators, and representatives on the associated date. Whether the president was democratic or republican onthe associated date. \
In snp data set, there are `r nrow(snp)` observations and `r ncol(snp)` variables. It includes the closing values of the S&P stock index on the associated date ranging from `r range(snp$year)[1]` to `r range(snp$year)[2]`.\
In unemployment data set, there are `r nrow(unemployment)` observations and `r ncol(unemployment)` variables. It indicates the percentage of unemployment every month ranging from `r range(unemployment$year)[1]` to `r range(unemployment$year)[2]`.

# Problem 2
```{r}
# import Mr. Trash wheel data set (skip figure)
trash_wheel <- read_excel("./202309 Trash Wheel Collection Data.xlsx", "Mr. Trash Wheel", "A2:N586")
# use reasonable variables
trash_wheel <- janitor::clean_names(trash_wheel)
trash_wheel <- trash_wheel |>
  dplyr::select(-year, -month) |>
  # add new homes_powered variable
  mutate(homes_powered = weight_tons * 500/30)
```

```{r}
# import Professor Trash wheel data set (skip figure)
pf_trash <- read_excel("./202309 Trash Wheel Collection Data.xlsx", "Professor Trash Wheel", "A2:M108")
# use reasonable variables
pf_trash <- janitor::clean_names(pf_trash)
pf_trash <- pf_trash |>
  dplyr::select(-year, -month) |>
  # add new homes_powered variable
  mutate(homes_powered = weight_tons * 500/30)
```

```{r}
# import Cwynnda Trash wheel data set (skip figure)
G_trash <- read_excel("./202309 Trash Wheel Collection Data.xlsx", "Gwynnda Trash Wheel", "A2:L157")
# use reasonable variables
G_trash <- janitor::clean_names(G_trash)
G_trash <- G_trash |>
  dplyr::select(-year, -month) |>
  # add new homes_powered variable
  mutate(homes_powered = weight_tons * 500/30)
```

```{r}
# add new variable to track trash wheel type
G_trash <- G_trash |>
  mutate(type = "Gwynnda Trash Wheel")
pf_trash <- pf_trash |>
  mutate(type = "Professor Trash Wheel")
trash_wheel <- trash_wheel |>
  mutate(type = "Mr. Trash Wheel")
# combine data
combined_trash <- bind_rows(trash_wheel, pf_trash, G_trash)
combined_trash <- combined_trash |>
  arrange(date)|>
  separate(date, c("year", "month", "day" ), "-")|>
  mutate(month= month.name[as.numeric(month)])
```
In Mr. Trash Wheel data set, there are `r nrow(trash_wheel)` observations and `r ncol(trash_wheel)` variables. In Professor Trash Wheel data set, there are `r nrow(pf_trash)` observations and `r ncol(pf_trash)` variables. In Gwynnda Trash Wheel data set, there are `r nrow(G_trash)` observations and `r ncol(G_trash)` variables.\
In combined data, there are `r nrow(combined_trash)` observations and `r ncol(combined_trash)` variables. Key variable `homes_powered` is number of homes powered by the trash, `type` is the source of trash wheel.

```{r}
# total weight of trash by Professor Trash Wheel
weight_trash <- sum(pull(pf_trash, weight_tons))
# total number of cigarette butts by Gwynnda in July 2021
G_July <- combined_trash |>
  filter(type=="Gwynnda Trash Wheel"& month=="July" & year=="2021")
cig <- sum(pull(G_July, cigarette_butts))
```
The total weight of trash collected by Professor Trash Wheel is `r weight_trash`.\
The total number of cigarette butts collected by Gwynnda in July of 2021 is `r cig`.


# Problem 3
```{r}

```
